Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-04-22 manuscript/manuscript.tex
addressed chloé's comments

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

\subsection{A case study: the consensus network}
\label{results:consensus}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{./figures/figure_3.pdf}
  \caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink.}
  \label{fig:consensus}
\end{figure}

@@

\subsection{Adjusting for instability leads to similar conclusions}

Most of the network methods, even the consensus, were highly unstable, raising questions about the veracity of the results. We built a consensus network using the 30 solutions from the 6 different methods on 5 different splits containing 80\% of the data (Section \ref{methods:comparison}). Such network would capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the settings (percentile 99\%). The resulting stablility-based consensus was composed of 68 genes. This network shares most of the proterties of the consensus: breast cancer susceptibility genes are overrepresented (P-value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P-values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively), the selected genes are more central than average (P-value = 1.1 \texttimes{} 10\textsuperscript{-14}), and 19 of them are isolated.

However, the shared global properties of the network, the lack of stability results in different genes being selected. In this case, the most central gene of the network is \emph{CUL3}, which is absent from the consensus network. Indeed, \emph{CUL3} has a low association score in both GENESIS and BCAC (P-values of 0.04 and 0.26, respectively). \emph{CUL3} is a component of Cullin-RING ubiquitin ligases. It impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the ability of GWAS to unravel the mechanisms leading to complex diseases has been put into questions \cite{boyle_expanded_2017}. The omnigenic model proposes that gene functions are interwoven with each other in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to the discovery an uninformative wide-spread pleiotropy. Other than such problems, discovery in GWAS is hindered by a conservative statistical framework. Network methods tackle both issues by using both the association score and an interaction networks to consider the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided biomarker discovery have been proposed in recent years. In this article we evaluate the relevance of six network methods by examining a GWAS dataset on familial breast cancer.

Most of these methods produced a relevant subset of biomarkers, recapitulating known breast cancer susceptibility genes. On median, the selected genes and SNPs were more central than average, in accordance with the observation that disease genes are more relatively central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting a given random pair of nodes, making them more likely to be selected by these network methods. Hence, further work is needed characterize the impact of centrality on these methods' outputs. Despite their similarities, the solutions were notably different. At one end of the spectrum, SConES and heinz preferred small, highly associated solutions, providing a conveniently short list of biomarkers, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provide biological hypotheses, they require further analyses, which might deter unexperienced practitioners. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Despite these differences, all solutions are equally bad at discriminating cases from controls. In addition, all methods are remarkably unstable, yielding different solutions for slightly different inputs. This might partly be caused by the instability of the P-values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P-values, which favors only the most extreme ones, leads to improved stability. Another source of instability might be the redundancy inherent to biological networks. We built a stable consensus, which accounts for that instability. Encouragingly, this network shared the same biological and topological properties as the individual solutions, despite the few genes in common due to instability. Overall, this supports that instability might be caused by redundant but equivalent biological mechanisms, and hence it validates the conclusions obtained on the individual solutions and the consensus.
\cazcom{}{I think we need to discuss this. Why aren't we able to classify cases from controls (multiple non-genetic factors, disease heterogeneity...)? Why is our approach still useful if we aren't, in fact, discovering biomarkers? (it points at relevant pathways, etc) }

To overcome the limitations of the individual methods while exploiting their strengths, we propose combining them into a consensus subnetwork. We use a straightforward aggregation, including any node that was recovered by at least two methods. The resulting network is a synthesis of the altered mechanisms: it is smaller than the largest solutions (HotNet2, SigMod and dmGWAS), which makes it more manageable, and includes the majority of the strongly associated smaller solutions (SConES and heinz). The consensus subnetwork captures mechanisms and genes known to be related to cancer, recovering known breast cancer susceptibility genes as well as genome regions associated to breast cancer susceptibility. However, thanks to its smaller size and its network structure, it provides compelling hypotheses of non-canonical mechanisms involved in carcinogenesis, like mitochondrial translation and chaperone activity.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.675\linewidth]{./figures/sfigure_4.pdf}
\caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). \textbf{(A)} Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink. This panel is identical to Figure~\ref{fig:consensus}. \textbf{(B)} Same network, but every gene name is indicated.}
\label{sfig:consensus_names}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figures/sfigure_6.pdf}
\caption{Biotypes of genes from the annotation that are not present in the HINT protein-protein interaction network.}
\label{sfig:biotypes_excluded}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=.85\linewidth]{./figures/sfigure_7.pdf}
\caption{Comparison of benchmark on high-throughput (HT) interactions to benchmark on both high-throughput and literature curated interactions (HT+LC). Grey lines represent no change in the statistic between the benchmarks (1 for ratios mean(HT) / mean(HT + LC), 0 for differences mean(HT) - mean(HT + LC)). \textbf{(A)} Ratios of the selected features between both benchmarks and of the active set. \textbf{(B)} Shifts in sensitivity and specificity. \textbf{(C)} Shift in Pearson correlation between benchmarks. \textbf{(D)} Ratio between the runtimes of the benchmarks. For gene network-based methods, inverted triangles represent the ratio of runtimes of the algorithms themselves, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2v2 took on average to compute the gene scores from SNP summary statistics. In general, adding additional interactions slightly improves the stability of the solution, but increases the solution size, has mixed effects on the sensitivity and specificity, and impacts negatively the required runtime of the algorithms.}
\label{sfig:lc_ht_comparison}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{./figures/sfigure_8.png}
  \caption{Overview of the solutions produced by the SConES on the GS and GM networks (Section \ref{methods:networks}) on the GENESIS dataset. \textbf{(A)} Manhattan plots of SNPs; in black,the method’s solution. The Bonferroni threshold (2.54 \texttimes{} 10\textsuperscript{-7}) is indicated by a red line. \textbf{(B)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs (SConES) or genes (other methods) in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(C)} Solution networks.}
  \label{sfig:scones_gsm}
\end{figure}

>>>>

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Yet, network methods are notably unstable, producing different results when the input data changes. Hence, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P-value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, a ubiquitin ligase related that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

\subsection{A case study: the consensus network}
\label{results:consensus}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{./figures/figure_3.pdf}
  \caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in Supplementary Figure \ref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

@@

\subsection{Adjusting for instability preserves global network properties}
\label{results:stable-consensus}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{./figures/figure_5.pdf}
  \caption{Stable consensus subnetwork on GENESIS (Section~\ref{results:stable-consensus}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the most central genes (\emph{CUL3}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in Supplementary Figure \ref{sfig:stable-consensus-names}.}
  \label{fig:stable-consensus}
\end{figure}

Most of the network methods, even the consensus, were highly unstable, raising questions about the reliability of the results. We built a new, stable consensus network using the 30 solutions obtained by running the 6 different methods on 5 different splits each containing 80\% of the data (Section \ref{methods:comparison}). Such a network is expected to capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the settings, corresponding to selecting 1\% of the genes. The resulting stablility-based consensus was composed of 68 genes (Figure \ref{fig:stable-consensus}). This network shares most of the proterties of the consensus: breast cancer susceptibility genes are overrepresented (P-value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P-values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes are more central than average (P-value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) are isolated.

However, although this new network exhibits similar global properties as the previous one, the lack of stability results in different genes being selected. In this case, the most central gene of the network is \emph{CUL3}, which is absent from the previous consensus network. Indeed, \emph{CUL3} has a low association score in both GENESIS and BCAC (P-values of 0.04 and 0.26, respectively). \emph{CUL3} is a component of Cullin-RING ubiquitin ligases. It impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the ability of GWAS to unravel the mechanisms leading to complex diseases has been called into question \cite{boyle_expanded_2017}. The omnigenic model proposes that gene functions are interwoven with each other in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to the discovery of an uninformative wide-spread pleiotropy. In addition, discovery in GWAS is hindered by a conservative statistical framework. Network methods tackle these two issues by using both the association score and an interaction network to take into consideration the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided biomarker discovery have been proposed in recent years. In this article we evaluate the relevance of six network methods by examining a GWAS dataset on familial breast cancer.

Most of these methods produced a relevant subset of biomarkers, recapitulating known breast cancer susceptibility genes. In general, the selected genes and SNPs were more central than average, in accordance with the observation that disease genes are more relatively central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting a given random pair of nodes, making them more likely to be selected by these network methods. Hence, further work is needed to characterize the impact of centrality on these methods' outputs. Despite their similarities, the solutions were notably different. At one end of the spectrum, SConES and heinz preferred small, highly associated solutions, providing a conveniently short list of biomarkers, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provide biological hypotheses, they require further analyses, which might deter unexperienced practitioners. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. To overcome the limitations of the individual methods while exploiting their strengths, we propose combining them into a consensus subnetwork. We use a straightforward aggregation, including any node that was recovered by at least two methods. The resulting network is a synthesis of the altered mechanisms: it is smaller than the largest solutions (HotNet2, SigMod and dmGWAS), which makes it more manageable, and includes the majority of the strongly associated smaller solutions (SConES and heinz). The consensus subnetwork captures mechanisms and genes known to be related to cancer, recovering known breast cancer susceptibility genes as well as genome regions associated to breast cancer susceptibility. However, thanks to its smaller size and its network structure, it provides compelling hypotheses of non-canonical mechanisms involved in carcinogenesis, like mitochondrial translation and chaperone activity.

Despite these differences, all solutions share two drawbacks. First, all methods are remarkably unstable, yielding different solutions for slightly different inputs. This might partly be caused by the instability of the P-values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P-values, which favors only the most extreme ones, leads to improved stability. Another source of instability might be the redundancy inherent to biological networks. We built a stable consensus, which accounts for that instability. Encouragingly, this network shared the same biological and topological properties as the individual solutions, despite the few genes in common due to instability. Overall, this supports that instability might be caused by redundant but equivalent biological mechanisms, and hence it validates the conclusions obtained on the individual solutions and the consensus. Second, all solutions are equally bad at discriminating cases from controls. Yet, the classification accuracy of network methods is similar to that of a classifier, which suggests that cases and controls are difficult to separate in this dataset. This might be due to unaccounted for environmental factors, and limited statistical power, which reduces the ability to identify relevant SNPs.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.675\linewidth]{./figures/sfigure_4.pdf}
\caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). \textbf{(A)} Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink. This panel is identical to Figure~\ref{fig:consensus}. \textbf{(B)} Same network, but every gene name is indicated.}
\label{sfig:consensus-names}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.675\linewidth]{./figures/sfigure_6.pdf}
  \caption{Stable consensus subnetwork on GENESIS (Section~\ref{results:stable-consensus}). \textbf{(A)} Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{CUL3}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink. This panel is identical to Figure~\ref{fig:stable-consensus}. \textbf{(B)} Same network, but every gene name is indicated.}
  \label{sfig:stable-consensus-names}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figures/sfigure_7.pdf}
\caption{Biotypes of genes from the annotation that are not present in the HINT protein-protein interaction network.}
\label{sfig:biotypes_excluded}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=.85\linewidth]{./figures/sfigure_8.pdf}
\caption{Comparison of benchmark on high-throughput (HT) interactions to benchmark on both high-throughput and literature curated interactions (HT+LC). Grey lines represent no change in the statistic between the benchmarks (1 for ratios mean(HT) / mean(HT + LC), 0 for differences mean(HT) - mean(HT + LC)). \textbf{(A)} Ratios of the selected features between both benchmarks and of the active set. \textbf{(B)} Shifts in sensitivity and specificity. \textbf{(C)} Shift in Pearson correlation between benchmarks. \textbf{(D)} Ratio between the runtimes of the benchmarks. For gene network-based methods, inverted triangles represent the ratio of runtimes of the algorithms themselves, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2v2 took on average to compute the gene scores from SNP summary statistics. In general, adding additional interactions slightly improves the stability of the solution, but increases the solution size, has mixed effects on the sensitivity and specificity, and impacts negatively the required runtime of the algorithms.}
\label{sfig:lc_ht_comparison}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{./figures/sfigure_9.png}
  \caption{Overview of the solutions produced by the SConES on the GS and GM networks (Section \ref{methods:networks}) on the GENESIS dataset. \textbf{(A)} Manhattan plots of SNPs; in black,the method’s solution. The Bonferroni threshold (2.54 \texttimes{} 10\textsuperscript{-7}) is indicated by a red line. \textbf{(B)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs (SConES) or genes (other methods) in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(C)} Solution networks.}
  \label{sfig:scones_gsm}
\end{figure}
